Stochastic Approximation Monte Carlo with Dynamic Update Factor (SAD)

### 2018-08-29 papers/histogram/sad.tex
update the paper and figure colors (ready for total edit!)

\subsection{Convergence}
{\color{red}
Although SAD determines the update factor ``dynamically'', the method must still 
satisfy the criteria for convergence. Similar to the discussion that was presented
on the convergence for SAMC, we approximate the total entropy change using an integral.
The update factor is integrated from the last time an energy has been discovered $t_L$ to
some minimum time $t_{\min}$ necessary for convergence.
\begin{align}
\Delta S_{\text{tot}} = \int_{t_L}^{t_{\min}} \gamma dt \approx t_L + t_L \ln\left(\frac{t_{\min}}{t_L}\right) \\
+ N_S t_L \Delta S_{\text{tot}} \left(\frac{1}{t_{\min}}-\frac{1}{t_L}\right) 
\end{align}

Simplifying, we see that as $t_{\min} \rightarrow \infty$ then $\gamma \rightarrow \infty$ satisfying
the requirement for SAMC convergence. We can solve for the minimum time
\begin{align}
\Delta S_{\text{tot}} = \Delta S_{\text{SAMC}} + N_S t_L \Delta S_{\text{tot}} \left(\frac{1}{t_{\min}}-\frac{1}{t_L}\right)
\end{align}
Solving for the minimum time necessary $t_{\min}$, we find that
\begin{align}
t_{\min} = t_L \left( \frac{N_S \Delta S_{\text{tot}}}{\Delta S_{\text{tot}} + \Delta S_{\text{SAMC}}} + 1\right)
\end{align}
Immediately we see that the minimum time necessary is roughly proportional
to the last time the simulation has found an energy times a ratio of entropies.

\jpsays{Are these right??? Check!}
We now consider the following limiting cases:
\subsubsection{$t \propto N_S \approx t_L$ and $ t << N_S$}
In this limiting case, we find that the update factor reduces to
\begin{align}
\gamma \approx \frac{\bar S}{N_S} \approx \frac{\bar S}{t}
\end{align}
Because $t \propto N_S$, the update factor scales as 1/t which is consistent
with the requirement for SAMC convergence.
\subsubsection{$t >> N_{S}^2$ and $ t\approx t_L$}
In this limiting case, we find that the update factor reduces to
\begin{align}
\gamma \approx \frac{N_S \bar S t_L}{t^2} \approx \frac{N_S \bar S}{t}
\end{align}
\subsubsection{$t >> t_L \bar S$}
In this limiting case, we find that the update factor reduces to
\begin{align}
\gamma \approx \frac{N_S}{t}
\end{align}

}

@@

For the square-well fluid, our simulations treat systems with a
well-width of $\lambda = 1.30$ and a filling fraction of
$\eta = 0.30$. We also explore various translation scales $\delta_0 = 0.005,
0.05,$ and $0.5$. The simulation explores the
energy space of the system at a reduced temperature of
$T_{\text{min}} = 0.2$ and $0.5$.  All simulations lead to the minimum important
energy $E_{\min}$ and maximum entropy energy $E_{\max}$
being calculated (with the exception of the WL methods where both
of these parameters are needed a priori).  In the sections below, we
examine systems parameters on the convergence of the methods.

We use the average entropy error vs moves a metric to compare
simulation runtimes and overall convergence. The overall accuracy
is determined by examining the fractional error of a particular method to
a `golden' standard. For each simulation, the `golden' standard is
SAMC for the best choice of $t_0$ for $10^{11}$ moves followed by TMMC.
\begin{align}
\epsilon_\text{avg} = 1 - \bigg\lvert\frac{S_\text{method}}{S_\text{golden}}\bigg\rvert
\end{align}
Simulations that take more iterations to reach the same entropy error as a
different method take longer to converge. This behavior is seen in
Fig.~\ref{fig:N50-ff0.3-avg-error}.

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-entropy-error-default.pdf}
  \caption{The average entropy error for each method is shown for $\lambda = 0.30$ with
  SAMC followed by TMMC serving as the `golden'
  standard.}\label{fig:N50-ff0.3-avg-error}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-entropy-error-slow.pdf}
  \caption{The average entropy error for each method is shown for $\lambda = 0.30$ with
  SAMC followed by TMMC serving as the `golden'
  standard.}\label{fig:N50-ff0.3-avg-error}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-entropy-error-fast.pdf}
  \caption{The average entropy error for each method is shown for $\lambda = 0.30$ with
  SAMC followed by TMMC serving as the `golden'
  standard.}\label{fig:N50-ff0.3-avg-error}
\end{figure}
\begin{figure}
\includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N500-entropy-error-default.pdf}
  \caption{The average entropy error is shown for $\lambda = 0.30$ with TMI acting as the
  `golden' standard.}\label{fig:n500-ff0.3}
\end{figure}

~\jpsays{We have discussed including a figure for the $\lambda = 30\%$ and $N = 50$
showing translation scale impact. Also a figure for the $\lambda = 30\%$ and $N = 50$
showing the impact of choice of minimum temperature.}

\subsection{A periodic system with $\lambda = 30\%$ and $N = 50$}

Figure~\ref{fig:N50-ff0.3-avg-error} shows the average entropy vs moves for
a system consisting of 50 atoms that are placed in a periodic simple cubic unit
cell with lattice constant chosen to give a $\lambda = 0.30$.  These particles
are randomly inserted into the system and random moves are proposed (with
a translation scale $\delta_0 = 0.05$).

\subsubsection{translation scale}
For the $\lambda = 30\%$ and $N = 50$ system, we explore the choice of $\delta_0$
as a parameter. Smaller translation scales should cause the given method to take
additional time to explore all energies.  A larger translation scale could result
in the simulation taking shorter or longer (because longer distance moves may end up
not being allowed). Ideal methods should scale linearly with the $\delta_0$, that is,
a decrease in the translation scale should result in a proportional increase in convergence
time. Some methods such as WL may not be able to handle arbitrary $\delta_0$.

\subsubsection{minimum temperature}
For the $\lambda = 30\%$ and $N = 50$ system, we explore the choice of $T_{\min}$
as a parameter. We have theoretically postulated that a dynamic version of SAMC which
we call SAD should depend upon this physically meaningful parameter. We thus explore
the behaviour of SAD for two different reduced temperatures.

\subsection{A periodic system with $\lambda = 30\%$ and $N = 500$}

A system is configured with dimensions $x = y = z = 19\sigma$.  A
filling fraction $\lambda = 0.30$ is chosen which yields $N = 500$
particles in a cubical lattice system.  These particles are randomly inserted
into the system and random moves are proposed (with a translation scale
$\delta_0 = 0.05$).

@@

A close examination of Fig.~\ref{fig:gamma-vs-t} shows that $\gamma$
for the Wang-Landau algorithm decreases much more rapidly than
$\nicefrac{1}{\sqrt{t}}$. It can be observed that all the methods from
this family struggle to converge if the update factor decreases faster
than $\nicefrac{10^5}{\sqrt{t}}$. Fig.~\ref{fig:n500-ff0.3} shows both
WL and SAMC with inappropriate choices of $t_0$ i.e. less than
$\nicefrac{10^5}{\sqrt{t}}$ failing to converge in a reasonable amount
of time. Thus, the update factor $\gamma$, shown as a function of
simulation moves in Fig.~\ref{fig:gamma-vs-t}, proves useful in
determining which methods will converge in a reasonable amount of time.
We compare each method to a `golden' standard which we define to be an
ideal choice of SAMC for $10^{11}$ moves followed by TMMC.  This gives an
unbiased comparison with our developed method SAD. The dynamic version
of SAMC which we call SAD performs as well as SAMC (assuming the user
can correctly identify an appropriate choice of $t_0$).  SAD
outperforms WL and WL-TMMC for the same given number of moves. The
absence of user defined parameters makes SAD an exciting alternative to
methods where user experience or successive implementation iteration
are needed to achieve reasonable results.

@@

We have introduced a new algorithm SAD (which is Stochastic Approximation
with the modification factor dynamically chosen) that effectively
samples the entire energy space for a variety of system sizes.  Also,
SAD proves effective in comparison with other Monte-Carlo methods at
converging to the correct density of states.  The ease of
implementation and the absence of user-defined parameters makes SAD an
ideal algorithm. SAD converges as quickly as SAMC (in the case of an ideally
chosen $t_0$ for our simulations. SAD doesn't suffer from the weakness of SAMC,
i.e. the user choosing parameters that could lead to the simulation not
converging in a reasonable amount of time. SAD also converges more quickly
than TMMC (the reader will remember that SAMC followed by TMMC is what each
method is compared against) and much more quickly than WL-TMMC.

>>>>

\subsection{Convergence}
Although SAD determines the update factor ``dynamically'', the method must still
satisfy the criteria for convergence. Similar to the discussion that was presented
on the convergence for SAMC, we approximate the total entropy change using an integral.
The update factor is integrated from the last time an energy has been discovered $t_L$ to
some minimum time $t_{\min}$ necessary for convergence.
\begin{align}
\Delta S_{\text{tot}} &= \int_{t_L}^{t_{\min}} \gamma dt \\
&\approx N_S \ln\left(\frac{t_{\min}}{t_L}\right)
   + \Delta S_{\text{tot}} \left(1-\frac{t_L}{t_{\min}}\right)
\end{align}
Simplifying, we see that as $t_{\min} \rightarrow \infty$ then $\gamma \rightarrow \infty$ satisfying
the requirement for SAMC convergence. We can solve for the minimum time
\begin{align}
\Delta S_{\text{tot}} \approx N_S \ln\left(\frac{t_{\min}}{t_L}\right)
+ \Delta S_{\text{tot}} \left(1-\frac{t_L}{t_{\min}}\right)
\end{align}
Solving for the minimum time necessary $t_{\min}$, we find that
\begin{align}
\frac{t_{\min}}{t_L} = e^{\frac{\langle S\rangle t_L}{t_{\min}}}
\end{align}
The solution for $t_{\min}$ can be written in terms of the Lambert-W function
\begin{align}
t_{\min} = \frac{3\langle S\rangle}{W(3\langle S\rangle e^{2\langle S\rangle})} t_L
\end{align}
For $\langle S\rangle \rightarrow 0$, we find that the minimum time
$t_{\min} \rightarrow t_L$ and for $\langle S\rangle \rightarrow \infty$,
we find that the minimum time $t_{\min} \rightarrow \frac{3}{2} t_L$.

We can examine the limiting cases for SAD and see that the behavior at
short times is analogous to the weight-based family algorithms $\gamma \approx 1$.
In the limit that the number of moves $t >> t_L$, $\gamma \approx \nicefrac{N_S}{t}$
which has a familiar 1/t form.

@@

For the square-well fluid, our simulations treat systems with a
well-width of $\lambda = 1.30$ and a filling fraction of
$\eta = 0.30$. We also explore various translation scales $\delta_0 = 0.005 and
0.05$. The simulations explore the
energy space of the systems at reduced temperatures of
$T_{\text{min}} = 0.2$ and $0.5$.  All simulations lead to the minimum important
energy $E_{\min}$ and maximum entropy energy $E_{\max}$
being calculated (with the exception of the WL methods where both
of these parameters are needed a priori).  In the sections below, we
examine the systems parameters impact on the convergence of the methods.

We use the average entropy error vs moves as a metric to compare
simulation runtimes and overall convergence. The overall accuracy
is determined by examining the fractional error of a particular method to
a `golden' standard. For each simulation, the `golden' standard is
SAMC for the best choice of $t_0$.
\begin{align}
\epsilon_\text{avg} = 1 - \bigg\lvert\frac{S_\text{method}}{S_\text{golden}}\bigg\rvert
\end{align}
Simulations that take more iterations to reach the same entropy error as a
different method take longer to converge. This behavior is seen in
Fig.~\ref{fig:N50-ff0.3-avg-error}.

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-entropy-error-default.pdf}
  \caption{The average entropy error for each MC method is shown for $\eta = 0.30$ and $T_{\min} = 0.2$ with
  SAMC $t_0 = 10^5$ serving as the `golden' standard.}\label{fig:N50-ff0.3-avg-error}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-entropy-error-slow.pdf}
  \caption{The average entropy error for each MC method is shown for $\eta = 0.30$ and $T_{\min} = 0.2$ with
  SAMC $t_0 = 10^5$ from Fig.~\ref{fig:N50-ff0.3-avg-error}
  serving as the `golden' standard.}\label{fig:N50-ff0.3-avg-error-slow}
\end{figure}
%\begin{figure}
  %\includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-entropy-error-fast.pdf}
  %\caption{The average entropy error for each method is shown for $\eta = 0.30$ with
  %SAMC followed by TMMC serving as the `golden'
  %standard.}\label{fig:N50-ff0.3-avg-error-fast}
%\end{figure}
\begin{figure}
\includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N500-entropy-error-default.pdf}
  \caption{The average entropy error for each MC method is shown for $\eta = 0.30$ and $T_{\min} = 0.5$ with
  SAMC $t_0 = 10^6$ serving as the `golden' standard.}\label{fig:n500-ff0.3}
\end{figure}

\subsection{A periodic system with $\eta = 30\%$ and $N = 50$}

Figure~\ref{fig:N50-ff0.3-avg-error} shows the average entropy vs moves
for a system consisting of 50 atoms that are placed in a periodic
simple cubic unit cell with lattice constant chosen to give a $\eta =
0.30$.  These particles are randomly inserted into the system and
random moves are proposed (with a translation scale $\delta_0 = 0.05$).
For the $\eta = 30\%$ and $N = 50$ system, we explore the choice of
$\delta_0$ as a parameter. Smaller translation scales cause the given
method to take additional time to explore all energies. Ideal methods
should scale linearly with the $\delta_0$, that is, a decrease in the
translation scale should result in a proportional increase in
convergence time. However, as can be seen from
Fig.~\ref{fig:N50-ff0.3-avg-error-slow} SAMC methods that rapidly
converged for $\delta = 0.05$ do not proportionaly converge for a
translation scale of $\delta = 0.005$. We see that SAD handles
arbitrary translation scales and proportionaly converges based on the
translation scale that is chosen.

\subsection{A periodic system with $\eta = 30\%$ and $N = 500$}

A system is configured with dimensions $x = y = z = 19\sigma$.  A
filling fraction $\eta = 0.30$ is chosen which yields $N = 500$
particles in a cubical lattice system.  These particles are randomly inserted
into the system and random moves are proposed (with a translation scale
$\delta_0 = 0.05$).

For the larger system, SAMC parameter values of $t_0$ that gave rapid convergence for
the $N = 50$ system with $\delta = 0.05$ are too small.  Several additional
larger $t_0$ parameters are chosen to compare with SAD.  SAD has significantly smaller
fluctuations in the average entropy error for a given number of moves.

@@

A close examination of Fig.~\ref{fig:gamma-vs-t} shows that $\gamma$
for the Wang-Landau algorithm decreases much more rapidly than
$\nicefrac{1}{\sqrt{t}}$. It can be observed  for the simulations that were carried out that all the methods from
the weight-based family struggle to converge if the update factor decreases too rapidlly. Fig.~\ref{fig:n500-ff0.3} shows
WL and SAMC with inappropriate choices of $t_0$ i.e. less than
$\nicefrac{10^5}{\sqrt{t}}$ failing to converge in a reasonable amount
of time. Thus, the update factor $\gamma$, shown as a function of
simulation moves in Fig.~\ref{fig:gamma-vs-t}, proves useful in
determining which methods will converge in a reasonable amount of time.
We compare each method to a `golden' standard SAMC simulation for an ideal choice of $t_0$.  This gives an
unbiased comparison with our developed method SAD. The dynamic version
of SAMC which we call SAD performs as well as SAMC (assuming the user
can correctly identify an appropriate choice of $t_0$ for SAMC).  SAD
outperforms WL and WL-TMMC for the same given number of moves. The
absence of user defined parameters makes SAD an exciting alternative to
methods where user experience or successive implementation iteration
are needed to achieve reasonable results.

@@

We have introduced a new algorithm SAD (Stochastic Approximation
with a Dynamic $\gamma$) that effectively
samples the entire energy space for a variety of system sizes.  Also,
SAD proves effective in comparison with other Monte-Carlo methods at
converging to the correct density of states.  The ease of
implementation and the absence of user-defined parameters makes SAD an
ideal algorithm. SAD converges as quickly as SAMC (in the case of an ideally
chosen $t_0$ for our simulations. SAD doesn't suffer from the weakness of SAMC,
i.e. the user choosing parameters that could lead to the simulation not
converging in a reasonable amount of time. SAD also converges more quickly
than Wl, TMMC, and WL-TMMC.
